---
title: "Loading Tapestri Data"
output:
  html_document:
    df_print: paged
    toc: true
editor_options: 
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Loading Tapestri Data}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r include=FALSE}

library(TapestriR)

# Packages needed for visualization 
library(tidyverse)

```

# Load data from various sources {.tabset}

Enable loading from various files. 

* H5 is the future format for multi-omics and multi-samples, supporting DNA and DNA + Protein analysis. 
* Loom supports only DNA mutation analysis.
* Insights export only supports DNA analysis.


## Load from Loom

Works only for DNA and single samples since loom files only contain 1 sample.  

```{r load_loom, cache=TRUE}

filename <- system.file("extdata", "PE11.cells.loom", package = "TapestriR")

variants = read_loom(filename,min_mutation_rate = 0.05)

filtered_variants = filter_variants(variants)
vaf=round(filtered_variants@data_layers$AD/filtered_variants@data_layers$DP, 3)
vaf[is.na(vaf)] <- 0
filtered_variants = add_data_layer(filtered_variants,'VAF',vaf)

filtered_variants
```

## Load multi-omics H5

Works for DNA and DNA + Protein data. Also works for multi-sample. Best practice is to create a multi-assay, multi-sample h5 in Pipeline, and apply filters before loading into R. 

Usage:
* Load multi-sample multi-omic h5 file
* Filter and normalize assays  


```{r load_h5}


filename <- system.file("extdata", "4_cell_line_mix_dna_protein.h5", package = "TapestriR")

experiment = read_tap(filename)

# Ideally, we would start by loading a filtered H5, but for now, we will load data with some basic filters.
filtered_variants = filter_variants(experiment$assays$dna)
vaf=round(filtered_variants@data_layers$AD/filtered_variants@data_layers$DP, 3)
vaf[is.na(vaf)] <- 0
filtered_variants = add_data_layer(filtered_variants,'VAF',vaf)

# Add the filtered data back to the experiment. This will be a subset of the rest of the assays to ensure that we have the same cells.
experiment = add_assay(moo = experiment,assay = filtered_variants, keep_common_cells = TRUE)

# Normalize using the clr method.
protein_counts_norm = experiment$assays$protein$data_layers$read_counts %>% clr_by_feature() %>% as_tibble(rownames = NA)
# Add normalized data to the protein assay.
experiment$assays$protein = add_data_layer(experiment$assays$protein,'normalized',protein_counts_norm)


normalized_dna_reads = normalize_dna_reads(experiment$assays$cnv$data_layers$read_counts)
experiment$assays$cnv = add_data_layer(experiment$assays$cnv,'normalized',normalized_dna_reads)


experiment
```

## Create your own multiomics object

Build a multiomics object from individual components instead of using the `read_tap` function. 

* Load variant assay and filter. 
* Load Protein assay and normalize.
* Load DNA read counts.
* Create a multiomics object and merge all 3 assays. This will ensure each assay has the same cells. 


```{r load_data}
ASSAY_NAME_VARIANT = 'dna'
ASSAY_NAME_PROTEIN = 'protein'

filename <- system.file("extdata", "4_cell_line_mix_dna_protein.h5", package = "TapestriR")

# Ideally, we would start by loading a filtered H5, but for now, we will send a filtering list to the function to mimic it.
variants = read_assay_h5(filename = filename, assay_name = ASSAY_NAME_VARIANT, min_mutation_rate = 0.005)

filtered_variants = filter_variants(variants)

vaf=round(filtered_variants@data_layers$AD/filtered_variants@data_layers$DP, 3)
vaf[is.na(vaf)] <- 0
filtered_variants = add_data_layer(filtered_variants,'VAF',vaf)

# Load the protein assay and normalize.
protein = read_assay_h5(filename = filename, assay_name = ASSAY_NAME_PROTEIN)
# Normalize using the clr method.
protein_counts_norm = protein@data_layers$read_counts %>% clr_by_feature() %>% as_tibble(rownames = NA)
# Add normalized data to the protein assay.
protein = add_data_layer(protein,'normalized',protein_counts_norm)

# Create a multiomics object (moo).
experiment = create_moo(experiment_name = basename(filename), cell_annotations = filtered_variants@cell_annotations)
experiment = add_assay(moo = experiment,assay = filtered_variants)

experiment = add_assay(experiment,protein, keep_common_cells = TRUE)

experiment
```

### X-Y plots

1) Select the proteins to plot on the X and Y axes.
2) Select a set of other feature(s) to color the plot by. If you choose more than one feature, each feature will be plotted in a subplot.

```{r fig.width = 12, fig.height = 8}


##################
# Select the proteins to plot on X and Y.
##################

# protein_x = 'CD34'
# protein_y = 'CD38'

protein_x = 1
protein_y = 2

##################
# Select 1 or more features to color by.
# color_by should be a vector of the column header you want to color by.
##################

# All proteins
color_by = experiment$assays$protein$data_layers$normalized

# Select a few proteins.
# color_by =  experiment$assays$protein$data_layers$normalized %>% select('CD110','CD117')

# Select a few variants.
# color_by =  experiment$assays$dna$data_layers$NGT %>% select(1:10) %>% mutate_all(as_factor) %>% mutate_all(recode_genotypes)


p  = tapestri_scatterplot(x = experiment$assays$protein$data_layers$normalized[[protein_x]], 
                 y= experiment$assays$protein$data_layers$normalized[[protein_y]], 
                 color_by = color_by)+ scale_colour_gradient2(low="yellow", mid='grey', high="darkred") 
p = p + xlab(protein_x) + ylab(protein_y) + ggtitle('Color by Protein Expression')
p

# Select a few variants.
color_by =  experiment$assays$dna$data_layers$NGT %>% select(1:10) %>% mutate_all(as_factor) %>% mutate_all(recode_genotypes)


p  = tapestri_scatterplot(x = experiment$assays$protein$data_layers$normalized[[protein_x]], 
                 y= experiment$assays$protein$data_layers$normalized[[protein_y]], 
                 color_by = color_by)
p = p + xlab(protein_x) + ylab(protein_y) + ggtitle('Color by Genotypes')
p

```


## Load the data from Tapestri Insights export

Works for DNA only. Export can be from a multi-sample project in Tapestri Insights. 

Usage:

* Extract the exported zip file.
* Read directory to create a new `Tapestri_Assay` object.

```{r}

export_dir <- system.file("extdata", "insights_2.2_export", package = "TapestriR")

# Show files
dir(export_dir)
variants = read_insights_export(export_dir = export_dir)

variants
```

