---
title: "Loading Tapestri Data"
output:
  html_document:
    df_print: paged
    toc: true
editor_options: 
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Loading Tapestri Data}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r include=FALSE}

library(TapestriR)

#packages needed for visualization 
library(tidyverse)
library(patchwork)


#packages needed for analysis
packages <- c("factoextra",  "NbClust", 'scran', 'uwot')

for(i in 1:length(packages)) {
  package = packages[i]
  if (!requireNamespace(package, quietly = TRUE))
    install.packages(package)

  require(package=package, character.only=TRUE)
}



ASSAY_NAME_VARIANT = 'dna'
ASSAY_NAME_READ_COUNT = 'cnv'
ASSAY_NAME_PROTEIN = 'protein'



```


### Load from Loom

Work for DNA only. Only single sample, since loom files only contain 1 sample.  

```{r load_loom, cache=TRUE}

filename <- system.file("extdata", "PE11.cells.loom", package = "TapestriR")

variants = read_loom(filename,min_mutation_rate = 0.005)

filtered_variants = filter_variants(variants)
vaf=round(filtered_variants@data_layers$AD/filtered_variants@data_layers$DP, 3)
vaf[is.na(vaf)] <- 0
filtered_variants = add_data_layer(filtered_variants,'VAF',vaf)


```

#### Demension Reduction

```{r}

#dimensional reduction using umap
set.seed(111)
umap_values <- umap(filtered_variants$data_layers$VAF, scale=TRUE, metric="manhattan", init="laplacian", pca=20) 

umap_layer = tibble(    
      x = umap_values[,1],
      y = umap_values[,2]
)

filtered_variants = add_analysis_layer(assay = filtered_variants, layer_name = 'umap_vaf',data = umap_layer)


```

#### Plot the data

```{r}

p = ggplot(filtered_variants@analysis_layers$umap_vaf)
p = p + geom_point(aes(x = x, y = y), alpha = 0.5, size=0.8)
p = p + theme_bw()
p

```


### Load multiomics H5

Works for DNA, DNA + Protein data. Also works from multisample. best practice is to create a multiassay, multisample h5 in pipeline, and apply filters before loading into R. 

usage:

1) Load variant assay and filter 
2) Load Protein assay and normalize
3) Load DNA read counts
4) create a multiomics object and merge all 3 assays. This will make sure each assay has the same cells. 


```{r load_h5}


filename <- system.file("extdata", "ABseq021.h5", package = "TapestriR")

# ideally would just start by loading filtered H5, but for now will send filtering list to function to mimic
variants = read_h5(filename = filename, assay_name = ASSAY_NAME_VARIANT, min_mutation_rate = 0.005)

filtered_variants = filter_variants(variants)

vaf=round(filtered_variants@data_layers$AD/filtered_variants@data_layers$DP, 3)
vaf[is.na(vaf)] <- 0
filtered_variants = add_data_layer(filtered_variants,'VAF',vaf)

### load protein assay and normalize
protein = read_h5(filename = filename, assay_name = ASSAY_NAME_PROTEIN)

# normalize using clr method
protein_counts_norm = protein@data_layers$read_counts %>% clr_by_feature() %>% as_tibble(rownames = NA)
# add normalized data to protein assay
protein = add_data_layer(protein,'normalized',protein_counts_norm)

### load cnv assay 
cnv = read_h5(filename,assay_name = ASSAY_NAME_READ_COUNT)
normalized_dna_reads = normalize_dna_reads(cnv$data_layers$read_counts)
cnv = add_data_layer(cnv,'normalized',normalized_dna_reads)

### create multiomics object (moo)
moo = create_moo(experiment_name = basename(filename), cell_annotations = filtered_variants@cell_annotations)
moo = add_assay(moo,filtered_variants)
moo = add_assay(moo,cnv,keep_common_cells = TRUE)
moo = add_assay(moo,protein, keep_common_cells = TRUE)

experiment = moo
```

#### X-Y plots

1) select which proteins to plot on X and Y axis
2) select set of other feature(s) to color the plot by. If you choose more then one feature, each feature will be plotted in a subplot

```{r fig.width = 12, fig.height = 8}


##################
# select the Proteins to plot on X and Y
##################

#protein_x = 'CD34'
#protein_y = 'CD38'

protein_x = 1
protein_y = 2

##################
# select 1 or more features to color by
# color_by should be a vector of column header you want to color by
##################

#all proteins
color_by = experiment$assays$protein$data_layers$normalized

#select a few proteins
#color_by =  experiment$assays$protein$data_layers$normalized %>% select('CD110','CD117')

#select a few variant
#color_by =  experiment$assays$dna$data_layers$NGT %>% select(1:10) %>% mutate_all(as_factor) %>% mutate_all(recode_genotypes)


p  = tapestri_scatterplot(x = experiment$assays$protein$data_layers$normalized[[protein_x]], 
                 y= experiment$assays$protein$data_layers$normalized[[protein_y]], 
                 color_by = color_by)+ scale_colour_gradient2(low="yellow", mid='grey', high="darkred") 
p = p + xlab(protein_x) + ylab(protein_y) + ggtitle('Color by Protein Expression')
p

#select a few variant
color_by =  experiment$assays$dna$data_layers$NGT %>% select(1:10) %>% mutate_all(as_factor) %>% mutate_all(recode_genotypes)


p  = tapestri_scatterplot(x = experiment$assays$protein$data_layers$normalized[[protein_x]], 
                 y= experiment$assays$protein$data_layers$normalized[[protein_y]], 
                 color_by = color_by)
p = p + xlab(protein_x) + ylab(protein_y) + ggtitle('Color by Genotypes')
p

```

#### Demension Reduction

```{r}

### dimensional reduction using umap
set.seed(111)
umap_values <- umap(experiment$assays$dna$data_layers$VAF, scale=TRUE, metric="manhattan", init="laplacian", pca=20) 

umap_layer = tibble(    
      x = umap_values[,1],
      y = umap_values[,2]
)

a = add_analysis_layer(assay = experiment$assays$dna, layer_name = 'umap_vaf', umap_layer)

experiment$assays$dna = add_analysis_layer(assay = experiment$assays$dna, layer_name = 'umap_vaf', umap_layer)

```


#### Plot the data

```{r}

p = ggplot(experiment$assays$dna$analysis_layers$umap_vaf)
p = p + geom_point(aes(x = x, y = y), alpha = 0.5, size=0.8)
p

```


### Load data from Tapestri Insights Export

Work for DNA only. Export can be from a multisample project in Tapestri Insights. 
usage:

1) extract exported zip file
2) read the `NGT.csv` and `Variants.csv` files
3) create new `Tapestri_Assay` object

```{r}

export_dir <- system.file("extdata", "insights_2.2_export", package = "TapestriR")

#show files
dir(export_dir)
variants = read_insights_export(export_dir = export_dir)


```

